\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-04-24 TeX/paper.tex

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={140mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{listings}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

@@

In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the U.S. Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al. \cite{Fowleretal2007}. The citation network we use consists of 9223 majority opinions written by the U.S. Supreme Court from 1954 to 2002. In this undirected network an edge occurs between two verticies (i.e. between two majority opinions) if one majority opinion cites the other. The endogenous statistics we include into this model are the number of edges, the edge-wise 0-shared partner (esp(0)) statistic and the dyad-wise 0-shared partner statistic (dsp(0)) (Snijders et al. \cite{SnijdersTomA.B..2006}, Hunter and Handcock \cite{Hunter.2006}). The esp(k) statistic counts the number of pairs which are directly linked and share exactly k partners. The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked. Mathematically written, esp(k) and dsp(k) are defined as 
$$\Gamma_{esp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm}g_{ij} \right)\right]$$
$$\Gamma_{dsp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm} \right)\right]$$
where $m\neq i$, $m \neq j$ and $\mathds{1}_k(x)$ is an indicator that is defined as $1$, if $k=x$ and $0$, if $k\neq x$.
({\it add explanation for adding esp(0) and dsp(0)}). 
The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases. In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$. The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times. The second row of matrix $C$, consists of the second actor's atribute, repeated $N$ times, and so on. Then, the statistics for a nodal covariate is defined as
$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute. An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$. In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$. Then, the corresponding statistic can be calculated the same way as nodal covariates. Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared componentwise.\\[0.3cm]
We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE. The MCMLE requires a sample size of at least $100000$ networks to converge. The bootstrapped MPLE was estimated by using 500 simulated networks. As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks. For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks. The results can be found in table ??.

>>>>

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={140mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage{listings}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

@@

In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the U.S. Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al. \cite{Fowleretal2007}. The citation network we use consists of 9223 majority opinions written by the U.S. Supreme Court from 1954 to 2002. In this undirected network an edge occurs between two verticies (i.e. between two majority opinions) if one majority opinion cites the other. The endogenous statistics we include into this model are the number of edges, the edge-wise 0-shared partner (esp(0)) statistic and the dyad-wise 0-shared partner statistic (dsp(0)) (Snijders et al. \cite{SnijdersTomA.B..2006}, Hunter and Handcock \cite{Hunter.2006}). The esp(k) statistic counts the number of pairs which are directly linked and share exactly k partners. The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked. Mathematically written, esp(k) and dsp(k) are defined as 
$$\Gamma_{esp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm}g_{ij} \right)\right]$$
$$\Gamma_{dsp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm} \right)\right]$$
where $m\neq i$, $m \neq j$ and $\mathds{1}_k(x)$ is an indicator that is defined as $1$, if $k=x$ and $0$, if $k\neq x$.
({\it add explanation for adding esp(0) and dsp(0)}). 
The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases. In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$. The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times. The second row of matrix $C$, consists of the second actor's atribute, repeated $N$ times, and so on. Then, the statistics for a nodal covariate is defined as
$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute. An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$. In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$. Then, the corresponding statistic can be calculated the same way as nodal covariates. Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared componentwise.\\[0.3cm]
We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE. The MCMLE requires a sample size of at least $100000$ networks to converge. The bootstrapped MPLE was estimated by using 500 simulated networks. As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks. For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks. The results can be found in table \ref{sc_results}.\\[0.3cm]

\begin{table}[]
\centering
\begin{tabular}{|>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|}
\hline
\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{l|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   \\ \cline{2-5} 
\multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St. Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St. Error \\ \hline \hline
Edges                                      &                                  &                                   & -5.657                           & 0.028                             \\ \hline
Nodecov.Salience                           &                                  &                                   & 0.592                            & 0.012                             \\ \hline
Nodecov.Year                               &                                  &                                   & 0.253                            & 0.012                             \\ \hline
Absdiff.Year                               &                                  &                                   & -2.470                           & 0.160                             \\ \hline
Absdiff.Year.Squared                       &                                  &                                   & 0.221                            & 0.066                             \\ \hline
DSP(0)                                     &                                  &                                   & 0.0005                           & 0.0002                            \\ \hline
ESP(0)                                     &                                  &                                   & -3.129                           & 0.012                             \\ \hline
\end{tabular}
\caption{Estimation results for the Supreme Court Network using the MCMLE and the bootstrapped MPLE}
\label{sc_results}
\end{table}
